---
title: "Ridge Regression and the Lasso"
author: "Reto Wüest"
date: "3/5/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data

We will use data from the 2009 British Social Attitudes Survey. You can download the data [here](bsas_short.RData).

Our response variable is:

- **imm_brit**: Out of every 100 people in Britain, how many do you think are immigrants from non-western countries?

We will use the following predictors:

- **resp_female**: Is the respondent female?
- **resp_age**: Age of the respondent
- **resp_household_size**: How many people live in respondent's household?
- **resp_party_cons**: Respondent is most likely to support the Conservative Party
- **resp_party_lab**: Respondent is most likely to support the Labour Party
- **resp_party_libdem**: Respondent is most likely to support the Liberal Democratic Party
- **resp_party_snp**: Respondent is most likely to support the Scottish National Party
- **resp_party_green**: Respondent is most likely to support the Green Party
- **resp_party_ukip**: Respondent is most likely to support the UK Independence Party
- **resp_party_bnp**: Respondent is most likely to support the British National Party
- **resp_party_other**: Respondent is most likely to support another party, no party, refused to say, or did not know
- **resp_newspaper**: Respondent normally reads a daily newspaper
- **resp_internet_hrs**: Hours the respondent spends using the internet (per week)
- **resp_religious**: Respondent regards himself/herself belonging to a particular religion
- **resp_time_current_employment**: Months respondent has spent with current employer
- **resp_urban_area**: Population density
- **resp_health**: Respondent's health
- **resp_household_income**: Total income of respondent's household

Let's have a look at the distribution of the response variable (the red dashed line shows the true percentage of non-Western immigrants):

```{r, echo=FALSE, message=FALSE}
library(foreign)
library(dplyr)
library(ggplot2)

load("bsas_short.RData")

response_count <- bsas_data %>%
  dplyr::group_by(imm_brit) %>%
  dplyr::summarise(no_imm_count = n()) %>%
  data.frame()

nw_immi <- 100 * (6.973 / 65.14)  # True percentage of non-western immigrants in the UK

bar_plot <- ggplot(data = response_count,
                   aes(x = imm_brit, y = no_imm_count)) +
  geom_vline(xintercept = nw_immi, colour = "red", linetype = "dashed") +
  geom_bar(stat = "identity", position = position_dodge()) +
  geom_text(aes(label = no_imm_count),
            position = position_dodge(width = 0.9), vjust = -0.5, size = 2) +
  scale_x_continuous(name = "Estimated percentage of non-Western immigrants") +
  scale_y_continuous(name = "Number of respondents") +
  theme_bw() +
  theme(axis.text.x = element_text(size = 10, margin = margin(t = 5, b = 5)),
        axis.text.y = element_text(size = 10, margin = margin(l = 5, r = 5)),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12))

bar_plot
```

The plot shows that 759 of 1049 respondents (or 72.4%) overestimate the percentage of non-Western immigrants in the UK.

## Data Preparation

```{r, echo=TRUE, message=FALSE}
library(foreign)
library(dplyr)

# Load data set
load("bsas_short.RData")

# Declare factor variables
bsas_data <- bsas_data %>%
  dplyr::mutate(resp_urban_area = factor(resp_urban_area,
                                         levels = 1:4,
                                         labels = c("rural", "rather rural",
                                                    "rather urban", "urban")),
                resp_health = factor(resp_health,
                                     levels = 1:4,
                                     labels = c("bad", "fair", "fairly good", "good")))
```

We will use the `glmnet()` function from the `glmnet` package to perform ridge regression and the Lasso. Before doing so, we use the `model.matrix()` function to create a matrix of predictors. This function automatically transforms any qualitative variables into dummy variables. This is important because `glmnet()` can only take quantitative inputs.

We remove the intercept from the matrix produced by `model.matrix()` because `glmnet()` will automatically include an intercept. We also exclude the predictor **resp_party_cons**, which will serve as the baseline.

```{r, echo=TRUE}
# Matrix of predictors (remove the intercept and resp_party_cons)
x <- model.matrix(imm_brit ~ . -1 -resp_party_cons, bsas_data)
```

## Ridge Regression

The `glmnet()` function has an `alpha` argument that determines what type of model is fit. If `alpha = 0`, then a ridge regression model is fit, and if `alpha = 1`, then a Lasso model is fit. We first fit a ridge regression model.

## The Lasso
